Parsing Postfix log files

### 2007-05-16 doc/logparser.tex
First attempt at describing timeout handling.  It'll need proofreading and editing for sure, but it's a start. Aborted delivery attempts section is written, but again will need work I'm sure. Minor cleanup on recent work.

\subsection{XXX}

@@

Postfix 2.2.x (and presumably previous versions) lacks explicit logging
when bounce notifications are generated; suddenly there will be log entries
for a mail which lacks an obvious source.  There are similarities to the
problem of child mails discussed in section
\ref{tracking-re-injected-mail}, but unlike the solution described therein
bounce notifications do not eventually have a log line which identifies
their source.  The solution is that heuristics must be used to identify
them, and those heuristics are:

@@

    \item The sender address will be <>.

    \item Neither \texttt{postfix/smtpd} nor \texttt{postfix/pickup} will
        have logged any messages associated with the mail, viz it was
        generated internally by Postfix, not accepted via SMTP or submitted
        locally by sendmail.

@@

    \item The sender address must be <>.

@@

@@

\subsubsection{Out of order log entries}

\label{out-of-order-log-entries}

>>>>

\subsection{Additional complications}

@@

Postfix 2.2.x (and presumably previous versions) lacks explicit logging
when bounce notifications are generated; suddenly there will be log entries
for a mail which lacks an obvious source.  There are similarities to the
problem of child mails discussed in section
\ref{tracking-re-injected-mail}, but unlike the solution described therein
bounce notifications do not eventually have a log line which identifies
their source.  Heuristics must be used to identify bounce notifications,
and those heuristics are:

@@

    \item The sender address will be $<>$.

    \item Neither smtpd nor pickup will have logged any messages associated
        with the mail, indicating it was generated internally by Postfix,
        not accepted via SMTP or submitted locally by sendmail.

@@

    \item The sender address must be $<>$.

@@

Some mail clients appear to send the following sequence of commands during
the SMTP session:

\begin{verbatim}
    EHLO client.hostname
    MAIL FROM: <sender@address>
    RCPT TO: <recipient@address>
    RSET
    MAIL FROM: <sender@address>
    RCPT TO: <recipient@address>
    DATA
\end{verbatim}

The client aborts the first delivery attempt after the first recipient is
accepted, then makes a second delivery attempt which it continues with
until the delivery is complete. XXX VERIFY THIS AND DETERMINE WHAT CLIENTS
EXHIBIT THIS BEHAVIOUR XXX

Once again Postfix does not log a message making the client's behaviour
clear, so once again heuristics are required to identify the when this
behaviour occurs.  In this case a list of all mails accepted during a
connection are saved in the connection state, and the accepted mails
examined when the disconnection line is parsed and the disconnection action
executed.  Each mail is checked for the following characteristics:

    \item Is the mail missing its cleanup log message?  Every mail which
        passes through Postfix will have a cleanup line; lack of a cleanup
        line is a sure sign the mail didn't make it too far.

    \item Were there exactly two smtpd log lines for the mail?  There
        should be a connection line and a mail accepted line.

If both checks are successful then the mail is assumed to be one of the
offending bogus mails and is discarded.  There will be no further entries
logged for such mails, so without identifying and discarding them they
accumulate in the state table and will cause clashes if the queueid is
reused.

@@

The DATA phase of the SMTP conversation is where the headers and body of the
mail are transferred.  Sometimes there is a timeout or the connection is
lost\footnote{For brevity's sake timeout will be used throughout this
section, but everything applies equally to lost connections.} during the
DATA phase; when this occurs Postfix will discard the mail and the parser
needs to discard the data associated with that mail.  It seems more
intuitive to save the data to the database, but if a timeout occurs there
won't yet have been any data gathered for the mail, so there is none
available to save; the timeout is saved with the connection data instead.

To deal properly with timeouts the parsing algorithm needs to do the
following in the timeout action:

    \item Record the timeout and associated data in the connection's
        results.

    \item If no mails have been accepted yet nothing needs to be done; the
        timeout action ends.  The timeout action is dependant on the clone
        action keeping a list of all mails accepted on each connection.

    \item Because of ESMTP pipelining the client can send MAIL FROM, RCPT
        TO and DATA commands in one packet, instead of sending each command
        individually and waiting for the reply before sending subsequent
        commands.  If pipelining is used and Postfix rejects the sender
        address or any of the recipient addresses there is no way for the
        client to tell which command was rejected, nor does there appear to
        be a way for the SMTP conversation to recover, so clients
        disconnect uncleanly.

        A timeout may thus apply either to an accepted mail, or a rejected
        mail.  To distinguish between the two cases the algorithm compares
        the timestamp of the last accepted mail against the timestamp of
        the last line logged by smtpd for that connection.  If the smtpd
        timestamp is later there was a rejection between the accepted mail
        and the timeout, therefore the timeout applies to a rejected mail;
        the timeout has already been recorded so the timeout action
        finishes.  If the mail acceptance timestamp is greater then the timeout
        applies to the just-accepted mail.

    \item The mail needs to be discarded, but there is yet another
        complication: sometimes the cleanup line is logged after the
        timeout line.  If the cleanup line has already been seen the mail
        can be discarded (both from global state and the connection's list
        of accepted mails) and the timeout action finishes.

        In the case where the cleanup line is still pending the algorithm
        updates a global cache of queueids, adding the queueid and the
        timestamp from the timeout line.  When the next cleanup line is
        parsed for that queueid the cache will be checked, and will be
        deemed part of the discarded mail and ignored if it meets the
        following requirements:

            \item The queueid must not have been reused yet, i.e. there
                isn't an entry in the state tables for the queueid.

            \item The timestamp of the cleanup line must be within six
                minutes of the mail acceptance timestamp.  Timeouts happen
                after five minutes, so six gives a one minute leeway.

        The next cleanup line must meet the criteria above for it to be
        discarded because not every connection where a timeout occurs will
        have a cleanup line logged for it; if the algorithm blindly
        discarded the next cleanup line it would in some cases be mistaken.
        Whether or not the line is discarded the queueid will be removed
        from the cache of timeout queueids.

\subsubsection{Pickup logging after cleanup}

\label{pickup-logging-after-cleanup}

Occasionally the pickup line logged when mail is submitted locally via
sendmail appears later in the logfile than the cleanup line for that mail.
This seems to occur during periods of particularly heavy load, so is most
likely due to process scheduling vagaries.  Normally if the queueid given
in the pickup line exists a warning is generated by the pickup action, but
if the following conditions are met it is assumed that the lines are out of
order:

    \item The only program which logged anything for the mail is cleanup.

    \item There is less than a five second difference in the timestamps of
        the cleanup and pickup lines.

As always with heuristics there may be circumstances in which these
heuristics incorrectly match, but none have been identified so far.
